---
title: "Philadelphia Weather Data Analysis"
author: "<a href=\"https://www.linkedin.com/in/fanli\">Fan Li</a>"
date: "February 22, 2015"
output: 
  html_document:
    theme: united
---

## Data Preparation

Randomly select 10 rows for examination.

```{r setup, message = FALSE, echo = FALSE}
# Load necessary R libraries and the dataset. 
library("knitr")
library("ggplot2")
library("lubridate")
library("scales")
library("RSQLite")
library("dplyr")

load("../data/weather.RData")  # See R/import_weather_data.R for data preparation

```


```{r sample_data, echo = FALSE}
weather %>%
  sample_n(10) %>%  # Sample n rows. Also see sample_frac()
  arrange(Year, Month, Day) %>%  # Sort dataframe by Year, Month and Day
  kable()  # Create a html table

```

The data files have seven columns^[http://learn.fi.edu/weather/data2/]:

* Columns 1, 2 and 3 contain the month, day, and year.
* Columns 4 and 5 are the daily high and low temperatures (in oF).
* Column 6 is the daily precipitation in inches (rain plus melted snow).^[Originally in 1/100 inches. Converted to inches for clarity]
* Column 7 is the daily snowfall, in inches.^[Originally in 1/10 inches. Converted to inches for clarity]


## Basic Operators

#### select(data, col.1, col.2, ...) 

Select/keep only the variables you mention. For example, if we are interested in the daily snowfall, we can create a subset of the data this way:

```{r}
# This is the same as weather[c("Year", "Month", "Day", "Snow")] but a little cleaner.
snowfall <- select(weather, Year, Month, Day, Snow)  
kable(head(snowfall))  # Create a html table

```

#### filter(data, condition.1, condition.2, ...) 

Because snowfall data prior to 1885 is missing, we will remove them from the analysis. We can also limit our analysis to Feburary as well.

```{r}
# This is the same as subset(snowfall, Year >=1885 & Month == 2)
snowfall <- filter(snowfall, Year >= 1885, Month == 2)
kable(head(snowfall))

```

#### arrange(data, col.1, col.2, ...)

Which Feburary day has the most snowfall in the history? We can have a list in reverse chronological order.

```{r}
snowfall <- arrange(snowfall, -Snow, -Year)
kable(head(snowfall))

```


#### group_by() + summarize()

Now, which Feburary is the most "snowy" month?

```{r}
feb.snow <- arrange(
                summarize(
                    group_by(snowfall, Year),
                    Snow.Sum = sum(Snow, na.rm = TRUE)
                ),
                -Snow.Sum, -Year
            )
kable(head(feb.snow))

```

So far the syntax difference from the core R may look trivial. However the dplyr way has several advantages:

* It appears cleaner and will give you more meaningful error message if anything goes wrong.
* Its standardized, SQL-like operators (verbs) makes data manipulation easier to think.
* Many core functions were written in C++ and have great speed gain versus standard R (to be discussed later).
* It can communicate with many data storage backends besides dataframe (to be discussed later).


## Pipe and Chaining

#### Pipe operator (%>%)

All above operators (and other operators in dplyr) take a dataframe as the first argument (input), followed by other parameters, and produce a dataframe as output.

The "%>%" operator was designed to "pipes" data or function output into the subsequent function and therefore convert nested structures into cleaner and more logical chain expressions.

```{r eval = FALSE}
data %>% function()  # equal to function(data)

foo() %>% bar()      # equal to bar(foo())

```

Therefore, the following chain can create the same "snowy" feburary dataset as shown above.

```{r}
feb.snow2 <- weather %>%
  select(Year, Month, Day, Snow) %>%                 # Step 1. Select relevant variables
  filter(Year >= 1885, Month == 2) %>%               # Step 2. Filter by year and month
  group_by(Year) %>%                                 # Step 3. Group by year
  summarize(Snow.Sum = sum(Snow, na.rm = TRUE)) %>%  # Step 4. Summarize monthly snowfall
  arrange(-Snow.Sum, -Year)                          # Step 5. Sort table by monthly snowfall/year

kable(head(feb.snow2))

```


#### Chaining

The chain operator also works with almost all other functions which takes a dataframe as the first argument. For example, we may pipe transformed data directly into ggplot to compare daily temperature between 1880 and 2000.

```{r, fig.width = 8, fig.height = 4, fig.align = "center"}
weather %>%
  filter(Year %in% c(1880, 2000)) %>%  # Data available from 1874
  mutate(Date = as.Date(ymd(paste(1234, Month, Day))),
         Year = as.factor(Year)) %>%
  select(Year, Date, HighTemp, LowTemp) %>%
  ggplot() +
  geom_linerange(aes(x = Date, ymax = HighTemp, ymin = LowTemp, color = Year), alpha = 0.25, na.rm = TRUE, size = 1) +
  scale_x_date(breaks = "1 month",  labels = date_format("%b")) +
  scale_y_continuous(name="Temperature (F)") +
  scale_color_manual(values = c("red", "blue")) +
  theme_bw() +
  theme(legend.justification=c(1,1), legend.position=c(1,1))


```


As another example, we can pipe the data into a lm model.




## Data Backends

#### Working with SQL database

As a trival example, we will convert our weather dataset into a RDB with RSQLite package. 

```{r}
# Create a database using RSQLite package
con <- dbConnect(SQLite(), "../data/weather.sqlite")

if(!"weather" %in% dbListTables(con))
  dbWriteTable(con, "weather", weather)

# Disconnect
dbDisconnect(con)

```

Next, we can directly working on this database through dplyr.

```{r}
weather.db <- src_sqlite("../data/weather.sqlite") %>%
  tbl("weather") %>%
  select(Year, Month, Day, Snow) %>%                 # Step 1. Select relevant variables
  filter(Year >= 1885, Month == 2) %>%               # Step 2. Filter by year and month
  group_by(Year) %>%                                 # Step 3. Group by year
  summarize(Snow.Sum = sum(Snow)) %>%  # Step 4. Summarize monthly snowfall
  arrange(-Snow.Sum, -Year)    

# Show result via collect()
kable(head(collect(weather.db)))

# Show SQL query
explain(weather.db)


```




#### Working with mutiple data sources

As a simple demonstration, we will compare the weather of Philadelphia and Boston in 2010. This is done by firstly merging Boston weather data^[http://www.wunderground.com/history/airport/KBOS/2015/2/23/DailyHistory.html] with our Philadelphia weather data.

`r kable(head(boston))`

```{r}
weather %>%
  transmute(Date = ymd(paste(Year, Month, Day)),
            Philly.High = HighTemp, Philly.Low = LowTemp) %>%
  inner_join(
    boston %>%
      transmute(Date = mdy(Date), Boston.High = HighTemp, Boston.Low = LowTemp),
    by = "Date"
  ) %>%
  head() %>%
  kable()

```








